Massively Parallel Monte Carlo Methods for Discrete Linear and Nonlinear Systems

### 2012-11-01 doc/prelim/presentation/prelim_presentation.tex
finsihed draft except for conclusion

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Parallelization of Monte Carlo methods for discrete systems
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Scalability is of concern
    \end{itemize}
  \item Development of a nonlinear solver leveraging Monte Carlo
    \begin{itemize}
    \item Application to nonlinear problems of interest
    \item Memory benefits
    \item Performance benefits
    \end{itemize}
  \end{itemize}
\end{frame}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Putting it All Together}

@@

  \begin{itemize}
  \item Used in practice everywhere
  \item Global reduction operations observed not to impede scalability
    \begin{itemize}
    \item Dot product
    \item Vector norms
    \end{itemize}
  \item Nearest neighbor computations have poor algorithmic strong
    scaling
    \begin{itemize}
    \item Matrix-vector multiply
    \item Weak scaling is better
    \end{itemize}
  \end{itemize}

@@

  \[
  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}
  \]
  \[
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}
  \]
\end{frame}

@@

  \pause
  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{direct_1.png}
    \end{center}
    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
        history per state, \sn{2.5}{3} total histories. 0.785 seconds
        CPU time.} }
  \end{figure}

@@

  \pause
  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
  \end{figure}

@@

  \begin{itemize}
  \item Work to date has used a physics-based formulation
    \begin{itemize}
    \item Radiation transport equations used as model system
    \item Transition probabilities built from problem-specific
      parameters
    \item Probabilities and weights must be rederived for each new
      equation set
    \end{itemize}
  \item Desire a generalization for all linear operator equations
    \begin{itemize}
    \item Requires a general parallel framework
    \item Requires implementation with a general linear algebra
      framework
    \item Operator, vector, and graph abstractions
    \end{itemize}
  \item Neumann-Ulam solvers and MCSA implemented using the Trilinos
    Petra frameworks
  \item Can be leveraged in modern physics implementations
  \end{itemize}

@@

@@

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{1.0}{ \input{msod_example.pdftex_t} }
    \end{center}
    \caption{\textbf{Overlapping domain example illustrating how
        domain overlap can reduce communication costs.}}
  \end{figure}

@@

@@

@@

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{1.0}{ \input{gentile_example.pdftex_t} }
    \end{center}
    \caption{\textbf{Gentile's example illustrating how domain
        decomposition can create reproducibility issues in Monte
        Carlo.}}
    \label{fig:gentile_example}
  \end{figure}

@@

@@

  \begin{itemize}
  \item CASL has leveraged the Drekar multiphysics code (SNL)
  \item Coupled fluid flow and heat transfer helps characterize many
    phenomena
  \item Drekar is massively parallel and leverages Newton-Krylov
    methods
  \item Propose using the largest Drekar problem to date as a
    challenge problem for the new Monte Carlo methods
  \end{itemize}

>>>>

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Parallelization of Monte Carlo methods for discrete systems
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Scalability is of concern
    \end{itemize}
  \item Development of a nonlinear solver for discrete systems
    leveraging Monte Carlo
    \begin{itemize}
    \item Application to nonlinear problems of interest
    \item Memory benefits
    \item Performance benefits
    \end{itemize}
  \end{itemize}
\end{frame}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Putting It All Together}

@@

  \begin{itemize}
  \item Widely used in practice
  \item Krylov methods require only the action of the operator
  \item Parallelism achieved through a handful of operations
  \item Global reduction operations observed not to impede scalability
    \begin{itemize}
    \item Dot product
    \item Vector norms
    \end{itemize}
  \item Nearest neighbor computations have poor algorithmic strong
    scaling
    \begin{itemize}
    \item Matrix-vector multiply
    \item Weak scaling is better
    \end{itemize}
  \item Short and long reccurrence relations available for
    orthogonalization
  \end{itemize}

@@

  \[
  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}
  \]
  \[
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}
  \]
\end{frame}

@@

  \begin{figure}[h!]
    \begin{center}
      \scalebox{1.0}{ \input{heat_eq_setup.pdftex_t} }
    \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. No thermal source was
      present.}}
  \end{figure}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{direct_1.png}
    \end{center}
    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
        history per state, \sn{2.5}{3} total histories. 0.785 seconds
        CPU time.} }
  \end{figure}

@@

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
  \end{figure}

@@

  \begin{itemize}
  \item Published work to date has used a physics-based formulation
    \begin{itemize}
    \item Radiation transport equations used as model system
    \item Transition probabilities built from problem-specific
      parameters
    \item Probabilities and weights must be rederived for each new
      equation set
    \end{itemize}
  \item Desire a generalization for all linear operator equations
    \begin{itemize}
    \item Requires a general parallel framework
    \item Requires implementation with a general linear algebra
      framework
    \item Operator, vector, and graph abstractions
    \end{itemize}
  \item Neumann-Ulam solvers and MCSA implemented using the Trilinos
    Petra frameworks
  \item Can be leveraged in modern physics implementations
  \end{itemize}

@@

  \begin{itemize}
  \item Each parallel process owns a piece of the domain
  \item Random walks must be transported across domains through
    communication
  \end{itemize}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Brunner's Work (2006 and 2009)}

    \begin{itemize}
    \item Looked at 4 communication patterns:
      \begin{itemize}
      \item Fully-locking synchronous
      \item asynchronous-send/synchronous-receive
      \item master/slave
      \item binary tree master/slave
      \end{itemize}
    \item Binary tree master/slave performed best but race conditions observed
    \item A later improvement to their work showed a fully asynchronous
      pattern to work best
    \item Poor scaling for unbalanced problems
    \end{itemize}

@@

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.75}{ \input{msod_example.pdftex_t} }
        \end{center}
        \caption{\textbf{Overlapping domain example illustrating how
            domain overlap can reduce communication costs.}}
      \end{figure}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Developed by Wagner and colleagues in 2010
      \item Each set contains the full domain
      \item Multiple sets replicate the domain
      \item Domains overlap within a set
      \item Reduces communication by a significant fraction
      \item Increases scalability through smaller processor sets
      \item Redundancy for resiliency (and useful work)
      \end{itemize}
    \end{column}

@@

  \begin{itemize}
  \item The amount of domain leakage dictates communication
  \item Per Siegel's 2012 work, define a leakage fraction
  \end{itemize}

  \[
  \lambda =
  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
    of\ \#\ of\ particles\ starting\ in\ local\ domain}
  \]

  \begin{itemize}
  \item Sigel observed $\lambda$ to be bound empirically to the mean
    free path of the system
  \item Optimum ratio of local to global domain size exists
  \item Experiments show feasibility for large-scale calculations
    \begin{itemize}
    \item Not bandwidth or latency limited in load-balanced case
    \end{itemize}
    \pause
  \item How do we define $\lambda$ empirically for linear operator
    equations? 
  \item Full-clip roullette for the MCSA correction?
  \end{itemize}

@@

  \begin{itemize}
  \item Procassini's 2005 work addressed some concerns
  \item Dynamic balancing replicates domains independently
  \item Linear operator equations may benefit - fixed source problems
  \end{itemize}

@@

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{0.7}{ \input{gentile_example.pdftex_t} }
    \end{center}
    \caption{\textbf{Gentile's example illustrating how domain
        decomposition can create reproducibility issues in Monte
        Carlo.}}
    \label{fig:gentile_example}
  \end{figure}

  \begin{itemize}
  \item 2005 work by Gentile outlines reproducability in parallel
    Monte carlo
  \item Maintaining random number sequences
  \item Floating point addition does not commute
  \item Useful for benchmarking and unit testing
  \end{itemize}

@@

Strategy
  \begin{itemize}
  \item Direct analogs between particle transport and Neumann-Ulam
    solvers 
  \item Aim for MSOD implementation and fully asynchronous
    communication patterns
  \end{itemize}

Questions
  \begin{itemize}
  \item How much domain overlap is suitable for linear operator
    equations? 
  \item Is memory a limitation for overlap and replication?
  \item Do the linear operator properties help select overlap?
  \item Will full-clip roullette perturb the MCSA solution?
  \item How much does MSOD facilitate scaling?
  \end{itemize}

@@

  \begin{itemize}
  \item CASL has utilized the Drekar multiphysics code (SNL)
  \item Coupled fluid flow and heat transfer helps characterize many
    phenomena
  \item Drekar is massively parallel and leverages Newton-Krylov
    methods
  \item Propose using the largest Drekar problem to date as a
    challenge problem for the new Monte Carlo methods
  \end{itemize}
